Tutorial: SQL-on-Hadoop Systems

### 2015-04-02 paper.tex
Intro and structure

\section{Introduction}
Enterprises are increasingly use Apache Hadoop, more specifically HDFS, as a central repository for all their data;
data coming from various sources, including operational systems, social media and the web, sensors and smart devices, as well as their applications.
At the same time many enterprise data management tools (e.g. from SAP ERP to Tableau) rely on SQL and many enterprise users are familiar and comfortable with it.
As a result, SQL processing over Hadoop data has gained significant traction over the recent years, and the number of systems that provide such capability has increased significantly.
There is a wide variety of solutions, system architectures and capabilities in this space.
This tutorial tries to shed a light on this matter.

%Over the last decade, the database field has witnessed significant major innovations and changes in enterprise data platforms. First came the wave of noSQL, which provide high scale-out systems, although at the expense of strong ACID transactions and declarative SQL processing. On the analytics side, Hadoop emerged as the platform for all analytics needs of the enterprise. Although Hadoop started with just MapReduce and HDFS, it evolved into a multi-framework system, which supports not only MapReduce but also Spark, Tez, SQL, and others. The driver that enables all these frameworks to work in a single cluster environment is the shared file system, Hadoop File System (HDFS). Given the popularity of SQL in the enterprise, it was soon evident that SQL processing over HDFS data is critical in this new emerging enterprise data platform.

In this tutorial, we use the term SQL-on-Hadoop to refer to systems that provide some level of declarative SQL(-like) processing over HDFS and noSQL data sources, using architectures that include computational or storage engines compatible with Apache Hadoop. We study the architecture of several SQL-on-Hadoop systems and classify them into five categories:

\vspace{-2mm}
\setcounter{mycounter}{0}
\begin{list}{\arabic{mycounter}.}{\usecounter{mycounter}\leftmargin=1.4em}
  \itemsep=0.5em
  \parsep=0.5em
  \parskip=0em
\item {{\bf Native systems with Big Data run-times:} These are systems that utilize a big data run-time such as Tez, or Spark to execute SQL queries. This category includes systems such as Hive and Spark SQL.}
\item{ {\bf Native systems with MPP architectures:} These are systems that employ an MPP database architecture that uses stand-alone long-running processing to execute SQL queries over HDFS data. This category includes systems such as Impala, Presto, Drill, and Big SQL}
\item{{\bf Database hybrids:} These are systems that combine a relational SQL engine with Hadoop run-time. This category includes systems such as Hadapt, HAWQ, and Vortex.}
\item{ {\bf Database connectors:} This category includes the extensions provided by major database vendors to reach out to Hadoop and ingest HDFS data.}
\item{ {\bf noSQL/Hbase specific variants:} These are systems that provide a declarative SQL layer on top of noSQL systems, HBase in particular. The systems in this category are Phoenix, and Splice Machine. Note that these systems focus on OLTP-like workloads, whereas all other categories address OLAP-like analytics}
\end{list}

An important aspect of SQL-on-Hadoop, which is orthogonal to all these categories, is the various file formats used to store the data. While text is an important format for in-situ SQL analysis, and ELT processing, various columnar formats, such as sequence, Avro, ORC, and Parquet, have been proposed to provide fast analytics query processing.

In this tutorial, we will first introduce these file formats, then we will discuss the characteristics, the pros and cons of each category in detail. In particular, we will examine their architectural differences, and cover in-depth their approach to indexing, updates, query optimization, as well as new algorithms they introduce in a 3-hour tutorial.

@@

\section{Category 3: Database hybrids}

>>>>

\section{Introduction}
Enterprises are increasingly using Apache Hadoop, more specifically HDFS, as a central repository for all their data;
data coming from various sources, including operational systems, social media and the web, sensors and smart devices, as well as their applications.
At the same time many enterprise data management tools (e.g. from SAP ERP and SAS to Tableau) rely on SQL and many enterprise users are familiar and comfortable with it.
As a result, SQL processing over Hadoop data has gained significant traction over the recent years, and the number of systems that provide such capability has increased significantly.
There is a wide variety of solutions, system architectures and capabilities in this space with each solution vendor claiming superiority of their solution.

In this tutorial we use the term SQL-on-Hadoop to refer to systems that provide some level of declarative SQL(-like) processing over HDFS and noSQL data sources, using architectures that include computational or storage engines compatible with Apache Hadoop. The purpose of this tutorial is to provide an overview of the SQL-on-Hadoop options and to try to categorize and compare those options.

\subsection{File formats}
Before starting presenting the various SQL-on-Hadoop architectures and systems, an important aspect of SQL-on-Hadoop, which is orthogonal to all these architecture, is the various file formats used to store the data. While text is an important format for in-situ SQL analysis, and ELT processing, various columnar formats have been proposed to provide fast analytics query processing. We present many of the available options and argue about their strengths.

\subsection{Categories}
For the main part of this tutorial, we study the architecture of several SQL-on-Hadoop systems and classify them into five categories:

\vspace{-2mm}
\setcounter{mycounter}{0}
\begin{list}{\arabic{mycounter}.}{\usecounter{mycounter}\leftmargin=1.4em}
  \itemsep=0.5em
  \parsep=0.5em
  \parskip=0em
\item {{\bf Native systems with Big Data run-times:} These are systems that utilize a big data run-time such as Tez \cite{tezsigmod15}, or Spark \cite{sparkhotcloud10} to execute SQL queries. This category includes systems such as Hive \cite{hive} and Spark SQL \cite{sparkSQL}.}
\item{ {\bf Native systems with MPP architectures:} These are systems that employ an MPP database architecture that uses stand-alone long-running processing to execute SQL queries over HDFS data. This category includes systems such as Impala \cite{Kornacker+15}, Presto, Drill \cite{drill}, and Big SQL \cite{bigsql}}
\item{{\bf Database hybrids:} These are systems that combine a relational SQL engine with Hadoop run-time. This category includes systems such as Hadapt, HAWQ, and Vortex.}
\item{ {\bf Database connectors:} This category includes the extensions provided by major database vendors to reach out to Hadoop and ingest HDFS data.}
\item{ {\bf noSQL/Hbase specific variants:} These are systems that provide a declarative SQL layer on top of noSQL systems, HBase in particular. The systems in this category are Phoenix, and Splice Machine. Note that these systems focus on OLTP-like workloads, whereas all other categories address OLAP-like analytics}
\end{list}

\section{Tutorial structure}

In this 3-hour tutorial, we will first present at a high-level the different architectures and categories for SQL-on-Hadoop processing and we argue about the pros and cons of each option.
Next, we present the file formats that are prevalent in storing data in Hadoop. We present each option and a comparison between them in terms of usability, data size (compression) and speed of processing. Then we move to the main part of this tutorial where we present each individual category in detail. In particular, for each category we discuss its characteristics, the pros and cons and present in more detail one or two representative systems.
We focus on the architectural differences, and cover in-depth their approach to indexing, updates, query optimization, as well as new algorithms they introduce.
After presenting each category and in the last part of the tutorial we discuss again and debate 
about architecture. That should be the engaging enough for the audience and act as a bridge to start a lively open discussion.

@@

\section{Category 3: Database hybrids}
The systems discussed above typically contain large amounts of
new code in order to create a SQL layer that is natively integrated
with Hadoop. Consequently many of the native systems are young and do
not have nearly as many production deployments as traditional database system
implementations in the analtyical database market, and are thus not as
well tested nor optimized for corner cases.

Database hybrids have emerged to fill this gap, leveraging
pre-existing robust database code for query processing and/or storage
in order to accerlate production-ready implementations of
SQL-on-Hadoop solutions. We will discuss three examples of database
hybrids: HadoopDB~\cite{hadoopdb} which uses large amounts of
PostgreSQL code; Hawq~\cite{hawq} which uses large amounts of
Greenplum code; and Vortex which uses large amounts of Actian
Vectorwise code.

We will, in particular, discuss the different architectures of these
hyrbid systems. In some cases, database files are stored in HDFS,
while in
other cases, database files are stored on the same physical machines
as HDFS, but on a separate file system. In some cases, data is
dynamically moved from Hadoop file formats to the native storage
structures of the DBMS~\cite{invisible-loading}.

Query execution, as well, is performed differently in different hyrbid
solutions. In some cases,
the entirety of 
query execution is performed by the database engine code, while in
other cases, query execution is split between database engine code and
native Hadoop execution engines such as MapReduce or Tez~\cite{split-execution}. We will also
discuss other integration points with the Hadoop stack, such as
integrations with Apache Hadoop YARN, and Apache HCatalog.
